Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-04 forecastflows.tex
start incorporating Lizie feedback

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple drivers of change, with a largely dominant influence of human activities \citep{Diaz2019}. This ongoing biodiversity crisis is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}. To support implementation of sustainable policies among the socioeconomic and environmental dimensions, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
Estimation of global biodiversity indicators and current trends depends on large-scale and long-term datasets---across terrestrial, freshwater, and marine ecosystems \citep[e.g.][]{Dornelas2018}. These data, gathered opportunistically and from multiple sources, are often unbalanced and have geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}. % ... lacks something

% Gap, problem: None of this is going well
The urgent need to answer policy-relevant questions has favored the proliferation of diverse methods developed by different researchers, lacking an overall coherence. Though there is no doubt nature is declining globally, significant uncertainty remains. There is no consensus on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}. Future projections also diverge considerably, due to a high model uncertainty at the ecological level. Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights \citep{Franklin2020}.

% Current workflows in ecology are not up to the task
% Here we introduce a universal workflow to adress this...
Current controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights. Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections. Here, we introduce an universal workflow that proposes to iteratively build upon all these steps, harmonizing both trend estimation and forecasting, with the aim of refocusing the debate on ecological questions and increasing the speed of scientific progress.

\vspace{0.5cm}
\noindent \textbf{Scientific method and workflows}
%  too much blah blah blah?

% General scientific method we all learn
Quantitative science rely on a model-based framework to confront hypothesis and data, making some approximations \citep{}. The general scientific method stresses out that the research question should guide both the design of the experiment and the corresponding model building. The experiment should then be conducted---according to this specific design. Finally, the resulting experimental data should be used to inform our model and differentiate between hypotheses---hopefully answering our initial question. However, this is often an idealized view.
% and does not reflect the complexity of ecology \citep{}. I already say it below

% Divergences from this are common 
Divergences from this scientific method are common. One explanation is the lack of rigor, leading to questionable practices, such as retrospectively crafting hypothesis to explain the results of a model rather than testing a clear pre-defined question (data-driven analysis, Fig. 1). But the reality of ecological research is also a major driver of the divergences from the ideal scientific method. Many important questions often cannot be addressed by conducting experiments and replications, and we must often rely on existing datasets to have a large-scale and long-term perspective \citep{Hilborn1997}. % I feel something is missing
However, this does not explain the persistent flaws and lack of overall coherence in ecological modeling. Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% Moving along the data-model space, step-by-step
% \item stressing the need to think about model before study design
% \item 
To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed to move along the data-model space, following a coherent sequence of steps (Fig 1). In particular, more efforts should be placed on model evaluation before incorporating any real data. This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure. Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations. The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models. At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty. This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations. It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % I think we would need to explain more this idea about parameters... of keep it for later

% More details on the workflow:  walk through the different steps
% the paragraph is a bit loooong
The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2). This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. The next step is to simulate data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
% Some issues might arise at this step
Structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable. This is the first feedback loop: if the model is not able to recover known parameter values, we must reconsider its structure. This way, we ensure the model works well on simulated data before incorporating real observations. We can then fit the model to real data, to obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. The next step is a new data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. This is a second feedback loop. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

\vspace{0.5cm}
\noindent \textbf{The workflow in practice}
% How to address current issues (two case studies)

@@

% Step back
%\item we need more data, and better question (relate this to both previous study cases)
%\item where can we best reduce uncertainties through new scientific insights?
%\item machine learning! If we change nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms! Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
%\item (ML has a way to collect and interpret large datasets...)
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data. 
In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
% If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...

>>>>

% Ecology has become super challenged to predict stuff for decision making
Nature is increasingly threatened by multiple, mainly human-driven, drivers of change \citep{Diaz2019}. The ongoing biodiversity loss is expected to increase in the next decades because of climate change, and will continue to alter ecosystem services and human well-being \citep{IPBES2019}. To support implementation of sustainable policies, it is critical to understand trends to date and be able to forecast future dynamics. 
% $\rightarrow$ Example of stuff to predict: population size and geographic distribution (terrestrial, freshwater, marine), extinction risk,...

% General way to do this so far... (bifurcated?)
In an era of rapidly expanding big data, robust models are essential both to estimate biodiversity trends and to forecast future changes.
Estimating global biodiversity trends depends on large-scale and long-term datasets \citep[e.g.][]{Dornelas2018}. These data, gathered opportunistically from multiple sources, are often unbalanced with massive geographic, temporal and taxonomic biases. Addressing these biases requires the use of appropriate statistical inference.
Forecasting future changes---under different plausible scenarios---generally relies on either correlative models or process-based models \citep{IPBES2019}. The latter, which focus on a mechanistic representation of ecosystem functioning, are often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022}.

% Gap, problem: None of this is going well
The current workflow has not led to consensus, and significant uncertainty remains---undermining the policy relevance of current results. % need a better topic sentence, as it is still a bit similar to topic sentence of the next paragraph?
There is no clear agreement on current species trends, with ongoing debates driven by widely varying reports that sometimes show conflicting trend directions \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}. Future projections also diverge considerably, due to a high model uncertainty at the ecological level. Predictive modeling is increasingly relying on overly complex models (with a huge number of parameters), making it less adequate to generate new scientific insights \citep{Franklin2020}.

% Current workflows in ecology are not up to the task
% Here we introduce a universal workflow to adress this...
Current controversies and focus on methodological aspects stem from the lack of coherence between current workflows in ecology \citep{Loreau2022, Talis2023, Johnson2024}. Each new model development is added as a separate layer, disconnected from the original research aim, the data stream and the previous scientific insights. Workflows should fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections. In particular, we should incorporate data simulation early in the workflow to identify flaws in model structure and constraints in data.
Here, we introduce a universal workflow that iteratively builds upon all these steps. We argue it would harmonize both trend estimation and forecasting, with the aim of refocusing the debate on ecological % and conservation?
questions and increasing the speed of scientific progress.

\vspace{0.5cm}
\noindent \textbf{Scientific method and workflows}

% General scientific method we all learn
Quantitative science relies on a model-based framework, to confront hypothesis and data, making some approximations \citep{}. The general scientific method stresses we should formulate a research question, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses.
However, this idealized scientific method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets to have a large-scale and long-term perspective \citep{Hilborn1997}.

% I can't remember precisely what we say about this paragraph
% Should it be the first paragraph of section 'The workflow in practice'? But I think it's nice to mention each part of the Fig. 1 here
This reality should drive researchers to use more robust and coherent methods, but this is rather leading to persistent flaws. 
Trend estimation still often relies on a one-way 'inference',  % need a better word? to stress out that there is no feedback, eah paper applied a slightly different model 
and most of the time is spent fitting the model to empirical data (Fig. 1). For forecasting, researchers focus on making predictions with complex models, but the steps of model building and parameterization are not very transparent and not clearly delineated (Fig. 1). The different parts of the model are often calibrated separately rather than as a whole, and some parameter values are just fixed based on experiments and expert knowledge.
% should we keep the sentence below
The lack of rigor can also lead to questionable practices---such as retrospectively crafting hypotheses to explain the results of a model rather than testing a clear predefined question (data-driven analysis, Fig. 1).
% mention the disconnection between trends and forecasts?

% There is thus room for improvment, with a workflow that addresses theses flaws while taking account the reality of working with ecological data.
% foreshadow stuf about simulation and feedback loops
To address these flaws while accounting for the realities of working with ecological data, a comprehensive workflow is needed that moves along the data-model space, in a coherent sequence of steps (Fig 1). In particular, more efforts should be placed on model evaluation before incorporating any real data. This would force the modeler to acknowledge that some parameters might be non-identifiable and to reconsider the model structure. Similarly, it is essential to assess whether model predictions---once parameters are informed by data---are consistent with observations. The strength of such workflow lies in its flexibility, making it applicable to a wide range of modeling approaches, from simple trend analyses to more complex process-based models. At each step, the modeler need to critically examine its understanding of ecological processes, questioning previous assumptions, and explicitly acknowledge sources of uncertainty. This approach has the potential to enhance model interpretability and allow for a more transparent evaluation of model strengths and limitations. It also replace parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting.

% More details on the workflow:  walk through the different steps
The first step of the workflow is to define an explicit research question and formulate some hypotheses (Fig 2). This involves making assumptions about the most influential drivers, within the specific context of our study. This step help us to think about the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and designed to answer our research question. The general idea is to start with a relatively simple model, that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. 
The next step is the first data simulation step. The idea is to craft some fake data that reflect our assumptions about the data structure and biases, by fixing parameter values to some values (which is straightforward if the parameters are interpretable). We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values. 
Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This lead to the second data simulation step, this time using our fitted model to generate predictions. This retrodictive check allows model output to be compared to observations. 
% adding something about parameters
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data: structural degeneracies in the model might become apparent, revealing that some parameters are highly non-identifiable. If the model fails to recover known parameter values, this flags % not sure about this word
the need to reconsider its structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

\vspace{0.5cm}
\noindent \textbf{The workflow in practice}

% How to address current issues (two case studies)
% Need an introduction here
Across the different fields of ecology---for both parameter estimation and forecasting---a systematic application of a coherent workflow holds the promise to highlight the opportunities to best reduce uncertainties through new scientific insights, toward the most critical steps. This will help refocus the debate on designing new hypothesis, formulating new questions---and guiding efforts to collect new data.

@@

% Step back
%\item we need more data, and better question (relate this to both previous study cases)
%\item where can we best reduce uncertainties through new scientific insights?
%\item machine learning! If we change nothing, what's the point of not doing ML? ML $>$ process-based without question, and ML $>$ trends without mechanisms! Where theory is lacking, or where we are far from mechanistic understanding, you might as well do ML!
%\item (ML has a way to collect and interpret large datasets...)

In a world where machine learning is rapidly advancing, there is no point of sticking to traditional methods if no changes are made. Machine learning may surpass process-based models if the latter lack a robust estimation of their parameters and fall in a complexity trap, at the cost of their interpretability. Similarly, trend analysis, when the focus is on methodological controversies (due to the lack of an iterative workflow) rather than on a robust mechanistic foundation, offers no clear advantage over machine learning.
% If we do not take care to apply this workflow, machine learning might be the more pragmatic approach...
